Creating optimal conditions for reproducible data analysis in R with `fertile`

### 2020-02-09 fertile.Rmd
added use case + start of lit review

# Extended Abstract

In the field of data science, an analyst has created *reproducible* work if: 1) their code and data files have been made publicly available, and 2) these files produce identical results when run by another analyst.

In the age of modern computer technology, where knowledge can be instantly shared worldwide, reproducibility is becoming incredibly important to the advancement of scientific knowledge.

The first reason for this is that reproducible research is transparent. When researchers provide the code and data used for their work in a well organized and reproducible format, readers are able to understand the steps taken to generate findings from raw data and determine whether or not they believe the results to be reliable or trustworthy. Without reproducibility, research results must be blindly trusted.

The second reason is that reproducibility allows for collaboration and extended research. The creators of reproducible research can easily receive feedback and recommendations on their work, and others interested in the research topic can see the code and easily apply the methods and ideas used in one project to their own future work with minimal effort.

However, while the necessity of reproducibility is clear, there are significant behavioral and technical challenges that impede its widespread implementation, and no clear consensus on standards of what constitutes reproducibility in published research.

In defining the steps toward creating reproducible analyses, different researchers place their emphasis on different areas, including file structure, documentation, the use of file paths in code, and accessibility. Within these categories, the recommendations for what steps to take in ensuring reproducibility also vary.

Much of the discussion around reproducibility is also generalized, written in a way applicable to users working with a variety of statistical software programs. Since all statistical software programs operate differently, generalized recommendations on reproducibility often cannot go in depth. While they provide some useful guidelines, they can often be relatively unhelpful, particularly to new analysts who are unsure how to apply such recommendations within the software programs they are using.

In order to be most effective, reproducibility recommendations must be tailored to specific software programs. However, there has been very little work on reproducibility that is tailored to this level of specificity. In this paper, attempt to remedy that, focusing on reproducibility in the R programming language with a concentration on the RStudio integrated development environment.

R is an ideal candidate for reproducibility recommendations due to the language's popularity for statistical analyses and the ease with which analysts can download and begin using the software.

We propose an R package called `fertile`, which focuses on a series of common reproducibility-harming mistakes programmers make while conducting data science projects in R, warning users of their errors and providing recommendations for how to correct them.

`fertile` operates in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that is already written for potential problems).

Retroactively, `fertile` is designed to be run on an R Project folder, analyzing the project structure and the analyst's use of file paths and considering randomness, among other areas of interest. `fertile` creates reproducibility reports, identifying mistakes that users have made and providing recommendations for remedies. For example, fertile might identify an analyst's use of randomness in code and recommend setting a seed.

Proactively, the package works similarly, identifying mistakes as they are made by the user and outputting an educational warning message identifying, and providing a solution for, the mistake. In its proactive warning system, `fertile` focuses primarily on the use of file paths. As users execute code, `fertile` looks for file paths passed to functions, identifying when absolute paths are provided or when paths point to a location outside of the project directory.

`fertile` is designed to educate the user about why the mistakes are problematic, and how to fix them. It is written for a wide audience, simple enough to be used by students in an introductory data science course.

@@

## Papers and Books

# Other articles

## Packages

>>>>

Data-based research cannot be fully *reproducible* unless the requisite code and data files produce identical results when run by another analyst.

As research is becoming increasingly data-driven, and because knowledge can be shared worldwide so rapidly, reproducibility is critical to the advancement of scientific knowledge.

When researchers provide the code and data used for their work in a well-organized and reproducible format, readers are more easily able to determine the veracity of any findings by following the steps from raw data to conclusions.

The creators of reproducible research can more easily receive more specific feedback (including bug fixes) on their work. Moreover, others interested in the research topic can use the code to apply the methods and ideas used in one project to their own work with minimal effort.

However, while the necessity of reproducibility is clear, there are significant behavioral and technical challenges that impede its widespread implementation, and no clear consensus on standards of what constitutes reproducibility in published research (@peng2009bio). Not only are the *components* of reproducible research up for discussion (e.g., need the software be open source?), but the corresponding *recommendations* for ensuring reproducibility also vary (e.g., should raw and processed data files be in separate directories?).

Much of the discussion around reproducibility is also generalized---it is written to be applicable to users working with a variety of statistical software programs. Since all statistical software programs operate differently, generalized recommendations on reproducibility are often shallow and unspecific. While they provide useful guidelines, they can often be difficult to implement, particularly to new analysts who are unsure how to apply such recommendations within the software programs they are using.

Thus, reproducibility recommendations tailored to specific software programs are more likely to be adopted.

In this paper, we focus on reproducibility in the R programming language with a concentration on projects that use the RStudio integrated development environment.

R is an ideal candidate for reproducibility recommendations due to the language's popularity for statistical analyses and the ease with which analysts can download and begin using the software.  
Several researchers and R users have recognized this and worked to publish a small body of papers and R packages focusing on increasing reproducibility in the R community.

Much of this work is narrowly tailored, with each package effectively addressising a small component of reproducibility--file structure, modularization of code, version control, etc. 
Many existing reproducibility packages, due to their focused nature, succeed at their area of focus well, but at a cost. They are often semi-complex to learn and operate, providing a barrier to entry for less-experienced data analysts.

There appear to be no existing tools that cater to analysts looking for an easy-to-learn, easy-to-implement, fast way to obtain a broad overview of their projects' reproducibility.
To address this, we present an R package called [`fertile`](https://github.com/baumer-lab/fertile) [^1]:{The authors gratefully acknowledge contributions from Hadley Wickham, Jenny Bryan, Greg Wilson, Edgar Ruiz, and other members of the `tidyverse` team.}, a low barrier-to-entry package which focuses on a series of common mistakes programmers make while conducting data science research in R.

## Academic Research

## Other Articles

## R Packages

# Methods

`fertile` operates in two modes: proactively (to prevent reproducibility mistakes from happening in the first place), and retroactively (analyzing code that has already been written for potential problems). 
`fertile` is available for download at [https://github.com/baumer-lab/fertile](https://github.com/baumer-lab/fertile)

## Proactive Use

Proactively, the package identifies potential mistakes as they are made by the user and outputs an informative message as well as a recommended solution. For example, `fertile` catches when a user passes a potentially problematic file path---such as an absolute path, or a path that points to a location outside of the project directory---to a variety of common input/output functions.

```{r setup, cache = F, include = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r, include = FALSE}
readr::write_csv(mtcars, "~/Desktop/my_data.csv")
library(fertile)
```

```{r, message = FALSE}
file.exists("~/Desktop/my_data.csv")
read.csv("~/Desktop/my_data.csv")
```

`fertile` is even more aggressive with functions (like `setwd()`) that are almost certain to break reproducibility---it causes them to throw errors.

```{r, message = FALSE}
setwd("~/Desktop")
```

The proactive features are activated immediately after loading the `fertile` package and require no additional effort by the user.

In addition to the interactive warning system, `fertile` provides several useful utility functions. Among other things, these include functions to check the type of a file and a way to create a copy of a project in a temporary directory.

```{r}
is_path_portable("~")

is_data_file("~/Desktop/my_data.csv")
```

## Retroactive Use

Retroactively, `fertile` analyzes potential obstacles to reproducibility in an RStudio Project (i.e., a directory that contains an `.Rproj` file), including the directory structure, the analyst's use of file paths, randomness, etc. `fertile}` creates reproducibility reports that identify potential mistakes and provide recommendations for remedies. For example, `fertile` might identify the use of randomness in code and recommend setting a seed.

Users can access the majority of `fertile`'s retroactive features through two primary functions.

The `proj_check()` function runs fifteen different reproducibility tests, noting which ones passed, which ones failed, the reason for failure, a recommended solution, and a guide to where to look for help. These tests include: looking for a clear build chain, checking to make sure the root level of the project is clear of clutter, confirming that there are no files present that are not being directly used by or created by the code, and looking for uses of randomness that to not have a call to `set.seed()` present. Subsets of the fifteen tests can be invoked using the select helper functions from `dplyr` and the `proj_check_some()` function.

```{r, eval = FALSE, include = FALSE}
knitr::write_bib("dplyr", file = "pkgs.bib")
```

```{r, message = FALSE}
library(tidyverse)
proj_check_some(".", contains("paths"))
```

The `proj_analyze()` function creates a report documenting the structure of a data analysis project. This report contains informations about all packages referenced in code, the files present in the directory and their types, suggestions for moving files to create a more organized structure, and a list of reproducibility-breaking file paths used in code.

## Sample Use Cases

`fertile` is designed in such a way as to be useful to users of all backgrounds and experiences. In this section, we provide two sample use cases demonstrating situations where users could benefit significantly from using the package.

### Introductory Data Science Student

Susan is taking an introductory data science course. This is her first time learning how to code, and she has not yet been exposed to ideas of research reproducibility. Her professor has assigned a data analysis project which must be completed in R Markdown. The project requires her to read in a data file located on her computer and use it to produce a graph.

She reads in the data, makes the graph, and knits her .Rmd file. It compiles successfully, so she submits the assignment. The next day, she receives an email from her professor saying that her assignment failed to compile and that she needs to make changes and try again.

Susan doesn't understand why it didn't work on the professor's computer when it did on her own. The professor recommends that she install `fertile` and run `proj_check()` on her assignment. She does this and gets a message informing her she used an absolute path to open her dataset, and she should use a relative path. She looks up what this means and then uses the new information to update her assignment. She resubmits and her second attempt is successful.

On future projects, she always loads and runs fertile to make sure her work is okay before submitting.

### 2. Experience R User

# Results

In an effort to understand the package's effectiveness, we also share preliminary results from a randomized, controlled experiment conducted on an undergraduate introductory data science course [^2]:{This study is approved by Smith College IRB, Protocol #19-032}. The purpose of the study is to determine whether `fertile` helps students produce data science research that is more likely to be reproducible.

# Conclusion

`fertile` is an R package that lowers barriers to reproducible data analysis projects in R. 
The features of `fertile` can be accessed almost effortlessly, making it easy for data analysts of all skill levels and backgrounds to gain a better understanding of how to make their work reproducible.
`fertile` is designed to educate the user about why the mistakes are problematic and how to fix them, promoting a greater understanding of reproducibility concepts in its users.
It is written for a wide audience, simple enough to be used by students in an introductory data science course, but still helpful to experienced analysts.
`fertile` also addresses a human challenge of reproducibility. In the moment, it often feels easiest to take a shortcut---to use an absolute path or change a working directory. However, when considering the long term path of a project, spending the extra time to improve reproducibility is worthwhile. `fertile`'s user-friendly features can help data analysts avoid these harmful shortcuts with minimal effort.

@@
